Ground state energy of quantum dots using the coupled cluster method

### 2018-06-11 project_2/article/article.tex
Review intermediates and part of RHF

        \subsubsection{Intermediate Computations}
        Looking closely at the amplitude equation in (\ref{eq:CCD_amp_1}) one might come
        to realize that this equation contains many of the same structures in
        several of the terms. This warrants the search for an algebraic transformation of the
        CCD amplitude equation that has the potential to reduce the amount of floating point
        operations needed to compute it. As it turns out, such terms exist and they will decrease
        the computing time necessary by an order of magnitude. We will define the following
        "intermediates",
        \begin{align}
            \label{eq:intermediate1}
            \chi^{ab}_{cd} &= \frac{1}{4}t^{ab}_{mn} u^{mn}_{cd} + \frac{1}{2}u^{ab}_{cd} \\
            \label{eq:intermediate2}
            \chi^n_j &= \frac{1}{2}t^{cd}_{jm} u^{mn}_{cd} \\
            \label{eq:intermediate3}
            \chi^a_d &= \frac{1}{2} t^{ac}_{nm} u^{nm}_{cd} \\
            \label{eq:intermediate4}
            \chi^{bm}_{jc} &= u^{bm}_{jc} + \frac{1}{2}t^{bd}_{jn}u^{mn}_{cd}
        \end{align}
        These intermediate structures will allow us to rewrite \autoref{eq:CCD_amp_1} to,
        \begin{equation}
        \begin{aligned}
            0 =& u^{ab}_{ij} + f^b_c t^{ac}_{ij}P(ab) - f^k_jt^{ab}_{ik}P(ij) \\
              &+ t^{cd}_{ij}\chi^{ab}_{cd} + t^{ab}_{in}\chi^n_jP(ij)
               - t^{bd}_{ij}\chi^a_dP(ab) \\
              &+ t^{ac}_{im}\chi^{bm}_{jc}P(ab)P(ij) + \frac{1}{2}t^{ab}_{im}u^{mn}_{jn}.
        \end{aligned}
        \end{equation}
        The importance of this "trick" will become apparent in due time.

    \subsection{Constructing the matrix elements}
        Having found the equations needed in order to find an estimate to the
        ground state energy using the coupled cluster doubles approximation is a
        well and dandy. But, we need basis functions to create the matrix
        elements needed to feed into the coupled cluster code.  Often these
        basis functions are not known and we have to use an approximation or
        utilize Hartree-Fock to create more optimized basis functions.

        \subsubsection{Harmonic oscillator basis}
            We will be looking at a system of two-dimensional quantum dots with
            a Coulomb repulsion.  If we assume, or make it so, that the
            repulsive two-body part is small we can use eigenfunctions of the
            one-body part our basis. In this case we have two-dimensional
            harmonic oscillator functions as eigenfunctions. We can then compute
            the matrix elements, $h_q^p$ and $u_{rs}^{pq}$, before feeding these
            into the coupled cluster code.

            In polar coordinates we can write the harmonic oscillator
            wavefunction for a single particle in two dimensions
            as\footnote{Note that this is without spin. As we are looking at
            fermions this means that each mode of the harmonic oscillator
            functions will be repeated twice.}.
            \begin{align}
                \phi_{nm}(r, \theta)
                &=
                N_{nm}
                (ar)^{\abs{m}}L_n^{\abs{m}}(a^2 r^2)e^{-a^2 r^2/2}
                e^{im\theta},
            \end{align}
            where $a = \sqrt{m\omega/\hbar}$ is the Bohr radius, $L_n^{\abs{m}}$
            is the associated Laguerre polynomials, $n$ and $m$ are the
            principal and azimuthal quantum numbers respectively and $N_{nm}$ is
            a normalization constant given by
            \begin{align}
                N_{nm}
                &= a\sqrt{\frac{n!}{\pi(n + \abs{m})!}}.
            \end{align}
            Included is also the spin, $\sigma$, of the wavefunction, which can
            be either up or down. This means each level, $(n, m)$, is doubly
            occupied. We also have that the wavefunctions are orthonormal
            \begin{align}
                \braket{n_1m_1\sigma_1}{n_2m_2\sigma_2}
                = \delta_{n_1n_2}\delta_{m_1m_2}\delta_{\sigma_1\sigma_2}.
            \end{align}
            The eigenenergy of a single harmonic oscillator is given by
            \begin{align}
                \epsilon_{nm} &= \hbar\omega(2n + \abs{m} + 1).
            \end{align}
            Our next job is now to create a mapping from the three quantum
            numbers $n$, $m$ and $\sigma$ to a single quantum number $\alpha$ as
            the matrices $h$ and $u$ use single indices for each wavefunction.
            In \autoref{fig:basis_states} we can see the energy levels that
            needs to be mapped.
            \begin{figure}
                \begin{center}
                    \begin{tikzpicture}[scale=0.8]
                        \begin{scope}
                            \foreach \i in {1, 2, 3} {
                                \draw(-1, \i - 1) node[anchor=east]
                                {$\epsilon = \i$};
                            }

@@

                            % Lowest energy level
                            \draw (3, 0) -- (5, 0);
                            \node at (3 + 0.75, 0) {$\uparrow$};
                            \node at (3 + 1.25, 0) {$\downarrow$};
                            \node[below, inner sep=.2cm] at (4, 0)
                            {$n = 0, m = 0$};
                        \end{scope}
                    \end{tikzpicture}
                \end{center}
                \caption{In this plot we can see the energy degeneracy of the
                lowest three energy levels in the two-dimensional quantum dot.
                Each arrow representes a spin up or a spin down state with the
                quantum numbers $n$ and $m$ as listed below.}
                \label{fig:basis_states}
            \end{figure}

            Starting from the bottom and working our way upwards from left to
            right we can label each line from $0$ to $n$ in increasing
            order. This enumeration will serve as our common quantum number
            $\alpha$\footnote{Note that we use greek letters $\alpha, \beta,
            \dots$ for the harmonic oscillator wavefunctions as opposed to latin
            letters for the general indices.}. We will only work with full
            \emph{shells}, i.e., we restrict our views to systems of $N$
            particles where $N$ will be a magic number which we get by counting
            all spin states for each energy level and the energy levels below.
            In \autoref{fig:basis_states} we can see the magic number $N \in [2,
            6, 12]$.

@@

            The two-body matrix elements are a little harder to work out as the
            harmonic oscillator wavefunctions are not eigenfunctions to the
            correlation operator. Luckily, from E. Anisimovas and A.  Matulis
            (Equation A2)\cite{anisimovas1998energy} we can get an analytical
            expression for the two-body matrix elements. We can then construct
            the antisymmetric two-body matrix elements in the harmonic
            oscillator basis.
            \begin{align}
                \bra{\alpha\beta}\ket{\gamma\delta}
                &= \braket{\alpha\beta}{\gamma\delta}
                - \braket{\alpha\beta}{\delta\gamma}.
            \end{align}
            As we only need to compute these once it is a good idea to save all
            non-zero values of $\bra{\alpha\beta}\ket{\gamma\delta}$ to a file.

@@

            The neatest, yet arguably the most abstract, way to write the Hartree-Fock
            equations for electron $i$\footnote{This is weird, as electron
            should be indistinguishable and therefore impossible to label.} is
            \begin{equation}
                f_i \varphi_i = \varepsilon_i \varphi_i,
            \end{equation}
            where $f_i$ is the Fock operator, $\varphi_i$ are eigenstates of the Fock
            operator consiting of a set of one-electorn wave funcitons called the
            Hartree-Fock molecular orbitals, and $\varepsilon_i$ are the eigenenergies
            of the Fock operator. The fock operator, in matrix notation, is given by
            \begin{equation}
                F_{pa} = H_{pq} +  J(D)_{pq} - \frac{1}{2}K(D)_{pq},
            \end{equation}
            where $h_i$ is the one-body operatore, while the two-body operator is
            divided into what we call a direct part,
            \begin{equation}
                J(D)_{pq} = \braket{pq}{rs}D_{sr},
            \end{equation}
            and an exchange part,
            \begin{equation}
                K(D)_{pq} = \braket{ps}{rq}D_{sr}.
            \end{equation}
            The direct part of the two-body operator is comparable to classical Coloumb
            repulsion, while the exchange does not have a classical analog as it arises
            from the antisymmetry requirement of the wavefunction. $D_{sr}$ is the
            density matrix of the system.

            Introducing a basis set transforms the Hartree-Fock equations into the Rothaan
            equations
            \begin{equation}
                FC = SC\varepsilon.
            \end{equation}
            This is a generalised eigenvalue problem where $S$ serves as an overlap matrix
            that must be there in case of non-orthogonal basis. Since the Fock matrix $F$
            depends on it's own solution through the orbitals, the eigenvalue problem must
            be solved iteratively\footnote{This is also the reason why the
            Hartree-Fock-Roothaan equations are often called the self-consistent-field
            procedure.}

            Because the RHF SCF method is a variational method, the initial
            wavefunction made up of the single-particle wavefunctions $\varphi_i$
            is varied until the minimal energy is found. The transformations necessary
            to make all these variations will now be contained in $C$. So if we start from
            a harmonic oscillator (HO) basis we can transform into a better basis, called a
            Hartree-Fock (HF) basis,
            \begin{equation}
                \ket{\Psi}^{\text{HF}}_p = \sum_{k=1}^L C_{kp} \ket{\Psi}_k^{\text{HO}}.
            \end{equation}
            We will see how the coupled cluster method with double excitations works with
            both the HO basis and the HF basis, but not equally well.

@@

@@

>>>>

        \subsubsection{Intermediate computations}
            Looking closely at the tensors in \autoref{eq:CCD_amp_1} one might
            come to realize that many of the terms share a common structure.
            This warrants the search for an algebraic transformation of the
            amplitude equation that has the potential to reduce the amount of
            floating point operations needed to compute it. As it turns out,
            such terms exist and they will decrease the computing time necessary
            by two orders of magnitude for the price of higher storage costs.
            We will define the following "intermediates"\footnote{Note that are
            more than one way of defining the intermediates.},
            \begin{align}
                \label{eq:intermediate1}
                \chi^{ab}_{cd}
                &= \frac{1}{4}t^{ab}_{mn} u^{mn}_{cd}
                + \frac{1}{2}u^{ab}_{cd} \\
                \label{eq:intermediate2}
                \chi^n_j &= \frac{1}{2}t^{cd}_{jm} u^{mn}_{cd} \\
                \label{eq:intermediate3}
                \chi^a_d &= \frac{1}{2} t^{ac}_{nm} u^{nm}_{cd} \\
                \label{eq:intermediate4}
                \chi^{bm}_{jc}
                &= u^{bm}_{jc} + \frac{1}{2}t^{bd}_{jn}u^{mn}_{cd}
            \end{align}
            See Appendix \ref{app:intermediates} for the derivation.  These
            intermediate structures will allow us to rewrite
            \autoref{eq:CCD_amp_1} to,
            \begin{equation}
                \begin{aligned}
                    0 &= u^{ab}_{ij} + f^b_c t^{ac}_{ij}P(ab)
                    - f^k_jt^{ab}_{ik}P(ij) \\
                    &\qquad
                    + t^{cd}_{ij}\chi^{ab}_{cd} + t^{ab}_{in}\chi^n_jP(ij)
                    - t^{bd}_{ij}\chi^a_dP(ab) \\
                    &\qquad
                    + t^{ac}_{im}\chi^{bm}_{jc}P(ab)P(ij)
                    + \frac{1}{2}t^{ab}_{im}u^{mn}_{jn}.
                \end{aligned}
            \end{equation}
            We will discuss the importance of doing this in
            \autoref{subsec:ccd_scaling}.

    \subsection{Constructing the matrix elements}
        Having found the equations needed in order to find an estimate to the
        ground state energy using the coupled cluster doubles approximation is
        all well and dandy. But, we need basis functions to create the matrix
        elements needed to feed into the coupled cluster code.  Often these
        basis functions are not known and we have to use an approximation or
        utilize Hartree-Fock to create more optimized basis functions.

        \subsubsection{Harmonic oscillator basis}
            We will be looking at a system of two-dimensional quantum dots with
            a Coulomb repulsion.  If we assume, or make it so, that the
            repulsive two-body part is small we can use the eigenfunctions of
            the one-body part as our basis of single particle functions. In this
            case we have two-dimensional harmonic oscillator functions as
            eigenfunctions. We can then pre-compute the matrix elements, $h_q^p$
            and $u_{rs}^{pq}$, before feeding these into the coupled cluster
            code.

            In polar coordinates we can write the harmonic oscillator
            wavefunction for a single particle in two dimensions
            as\footnote{Note that this is without spin. As we are looking at
            fermions this means that each mode of the harmonic oscillator
            function will be repeated twice.}.
            \begin{align}
                \phi_{nm}(r, \theta)
                &=
                N_{nm}
                (ar)^{\abs{m}}L_n^{\abs{m}}(a^2 r^2)e^{-a^2 r^2/2}
                e^{im\theta},
            \end{align}
            where $a = \sqrt{m\omega/\hbar}$ is the Bohr radius, $L_n^{\abs{m}}$
            is the associated Laguerre polynomials, $n$ and $m$ are the
            principal and azimuthal quantum numbers respectively and $N_{nm}$ is
            a normalization constant given by
            \begin{align}
                N_{nm}
                &= a\sqrt{\frac{n!}{\pi(n + \abs{m})!}}.
            \end{align}
            Included is also the spin, $\sigma$, of the wavefunction, which can
            be either up or down. This means each level, $(n, m)$, is doubly
            occupied. We also have that the wavefunctions are orthonormal
            \begin{align}
                \braket{n_1m_1\sigma_1}{n_2m_2\sigma_2}
                = \delta_{n_1n_2}\delta_{m_1m_2}\delta_{\sigma_1\sigma_2}.
            \end{align}
            The eigenenergy of a single harmonic oscillator is given by
            \begin{align}
                \epsilon_{nm} &= \hbar\omega(2n + \abs{m} + 1).
            \end{align}
            Our next job is now to create a mapping from the three quantum
            numbers $n$, $m$ and $\sigma$ to a single quantum number $\alpha$ as
            the matrices $h$ and $u$ use single indices for each wavefunction.
            In \autoref{fig:basis_states} we can see the lowest three energy
            levels that needs to be mapped.
            \begin{figure}
                \begin{center}
                    \begin{tikzpicture}[scale=0.8]
                        \begin{scope}
                            \foreach \i in {1, 2, 3} {
                                \draw(-1, \i - 1) node[anchor=east]
                                {$\epsilon = \i$};
                            }

@@

                            % Lowest energy level
                            \draw (3, 0) -- (5, 0);
                            \node at (3 + 0.75, 0) {$\uparrow$};
                            \node at (3 + 1.25, 0) {$\downarrow$};
                            \node[below, inner sep=.2cm] at (4, 0)
                            {$n = 0, m = 0$};
                        \end{scope}
                    \end{tikzpicture}
                \end{center}
                \caption{In this plot we can see the energy degeneracy of the
                lowest three energy levels in the two-dimensional quantum dot.
                Each arrow representes a spin up or a spin down state with the
                quantum numbers $n$ and $m$ as listed below. This pattern goes
                on indefinitly with the addition of one bar (two oscillators)
                per level.}
                \label{fig:basis_states}
            \end{figure}

            Starting from the bottom and working our way upwards from left to
            right we can label each line from $0$ to $n$ in increasing
            order. This enumeration will serve as our common quantum number
            $\alpha$\footnote{Note that we use greek letters $\alpha, \beta,
            \dots$ for the harmonic oscillator wavefunctions as opposed to latin
            letters for the general indices.}. We will only work with full
            \emph{shells}, i.e., we restrict our views to systems of $N$
            particles where $N$ will be a magic number which we get by counting
            all spin states for each energy level and the energy levels below.
            In \autoref{fig:basis_states} we can see the magic numbers $N \in [2,
            6, 12]$ by counting the levels in increasing order.

@@

            The two-body matrix elements are a little harder to work out as the
            harmonic oscillator wavefunctions are not eigenfunctions to the
            correlation operator. Luckily, from E. Anisimovas and A.  Matulis
            (Equation A2)\cite{anisimovas1998energy} we can get an analytical
            expression for the two-body matrix elements. We can then construct
            the antisymmetric two-body matrix elements in the harmonic
            oscillator basis.
            \begin{align}
                u^{\alpha\beta}_{\gamma\delta}
                &= \bra{\alpha\beta}u\ket{\gamma\delta}
                - \bra{\alpha\beta}u\ket{\delta\gamma}.
            \end{align}
            Computing the Coulomb elements is an extremely costly operation and
            we will thus only compute the non-zero orbital integrals, i.e., neglecting
            spin and antisymmetry, before saving these to a file which we can
            read several times.

@@

            The neatest, yet arguably the most abstract, way to write the
            Hartree-Fock equations for electron $i$ is
            \begin{equation}
                f \varphi_p = \varepsilon_p \varphi_p,
            \end{equation}
            where $f$ is the Fock operator, $\varphi_p$ are eigenstates of the
            Fock operator consisting of a set of one-electron wave functions
            called the Hartree-Fock molecular orbitals, and $\varepsilon_p$ are
            the eigenenergies of the Fock operator. The Fock operator, in matrix
            notation, is given by
            \begin{equation}
                f^p_q = h^p_q + J(D)^p_q - \half K(D)^p_q,
                %F_{pq} = H_{pq} +  J(D)_{pq} - \frac{1}{2}K(D)_{pq},
            \end{equation}
            where $h^p_q$ is the one-body operator, while the two-body operator
            is divided into what we call a direct part,
            \begin{equation}
                J(D)^p_q = \bra{pr}u\ket{qs}D_{sr},
            \end{equation}
            and an exchange part,
            \begin{equation}
                K(D)^p_q = \bra{pr}u\ket{sq}D_{sr}.
            \end{equation}
            Pay special attention to the non-antisymmetric two-body elements in
            the direct and exchange part.  The direct part of the two-body
            operator is comparable to classical Coloumb repulsion, while the
            exchange does not have a classical analog as it arises from the
            antisymmetry requirement of the wavefunction. The matrix $D_{sr}$ is
            the density matrix of the system and will be defined shortly.

            Introducing a basis set transforms the Hartree-Fock equations into
            the Rothaan equations
            \begin{equation}
                FC = SC\varepsilon.
            \end{equation}
            This is a generalised eigenvalue problem where
            \begin{align}
                S^{p}_{q} = \braket{p}{q},
            \end{align}
            i.e., the \emph{overlap matrix}. We now define the density matrix
            $D$ by
            \begin{align}
                D^{p}_{q} \equiv C^p_{\alpha}{C^{\alpha}_q}^{*},
            \end{align}
            Since the Fock matrix $F$ depends on it's own solution through the
            orbitals, the eigenvalue problem must be solved
            iteratively\footnote{This is also the reason why the Roothan-Hall
            equations are often called the self-consistent-field procedure.}

            Because the RHF SCF method is a variational method, the initial
            wavefunction made up of the single-particle wavefunctions
            $\varphi_i$ is varied until the minimal energy is found.
            \begin{align}
                E_{\text{RHF}}
                &= \para{
                    f^p_q - \half J(D)^p_q + \frac{1}{4}K(D)^p_q
                }D^q_p.
            \end{align}

            The transformations necessary to make all these variations will now
            be contained in $C$. So if we start from a harmonic oscillator (HO)
            basis, $\ket{\alpha}$ we can transform into a better basis, called a
            Hartree-Fock (HF) basis, $\ket{p}$.
            \begin{equation}
                \ket{p} = C^p_\alpha\ket{\alpha}.
            \end{equation}
            We will see how the coupled cluster method with double excitations
            works with both the HO basis and the HF basis, but not equally well.

@@

    \subsection{Scaling of the coupled cluster doubles method}
        \label{subsec:ccd_scaling}

@@

    \label{app:intermediates}
